\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-25 doc/incertae_sedis.tex
minor

The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to build
a comprehensive supertree \citep[see][]{redelings2017supertree}
covering all of life.
The approach is to combine information in
published phylogenies with a comprehensive taxonomy that supplies
taxon names.
The supertree method has been used with the Open Tree of
Taxonomy \citep[OTT hereafter, see][]{rees2017automated}, but in
principle is general.
The taxonomy is a crucial input for several
reasons: it covers a wider range of species, its list of names allows
for alignment of different phylogenetic estimates to each other, and
it provides names for clades.
Only a small proportion of known species
have been included in a phylogenetic analysis, thus a taxonomy is
important for achieving comprehensive coverage of known taxa.
Phylogenetic estimates collected from the published literature often
use different names for the same species.
Lists of synonyms in OTT
(and other taxonomies) allow data curators to ``align'' input
phylogenies to the taxonomy \citep[see][ for discussion of the
curation process in the Open Tree of Life project]{McTavishEtAt2015}.
This alignment is important for recognizing when two different
estimates refer to the same taxon.
Biologists are familiar with names
for ``higher'' taxa (taxa above the species rank).
Thus, naming clades
in the supertree makes the tree easier to navigate and use.

While some taxonomies may indicate hybrid taxa, the supertree approach
described here does not consider any hybridization.
Thus the taxonomic
hierarchy can be converted to a tree.
While not all taxonomists
believe that named taxa should correspond to clades \citep[see, for
example,][]{HorandlS2010}, the priniciples of phylogenetic
classification are used by such a large majority of practicing
taxonomists, that we have chosen to treat the taxonomic tree that
mirrors the hierarchy of the taxonomy as an estimate of phylogenetic
relationships.
We refer to labels of taxonomy nodes as ``taxon
names'', and assume that taxon names and taxonomy nodes have a one-to-
one correspondence.
Note that the taxonomy may also contain out-degree
1 nodes.
These nodes correspond to monotypic taxa which contain a
single child of lower taxonomic rank.

@@

Here describe a supertree method so that we can resolve taxonomic
uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa.
Developing this method required deciding on a set of
operational semantics to be used when interpreting the \emph{incertae
sedis} annotation.
These semantics affect the interpretation of the
phylogenetic statements being made by the input taxonomy and the rules
for applying taxonomic names to the final supertree.
In addition to
discussing how to interpret the \emph{incertae sedis} annotation,
we describe the algorithmic changes to the supertree pipeline that
were required to adequately represent the taxonomic uncertainty.
The
approach described here allows us to include thousands of new taxa in
our supertree analysis that were previously filtered out.
It will also
enable the Open Tree of Life project to include extinct taxa, since
many of these taxa are \emph{incertae sedis}.
This makes substantial
progress towards our goal of comprehensive inclusion of known taxa
into the Open Tree of Life summary tree.

@@

The supertree algorithm of \citet{redelings2017supertree} takes as
input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of leaf-labeled rooted input phylogenies and a single rooted taxonomy
tree $\taxonomy_{p}$ that is ranked below all of the input trees.
As
discussed in \citet{redelings2017supertree}, the preferred tree would
display as many of the highly ranked input splits as possible.
If
computational barriers were not an issue, the method would act by
first producing a ranked list of rooted splits:
$\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$ where
$S(T_{i})$ denotes a list of non-trivial rooted splits created from a
post-order traversal of tree $i$ and '$+$' denotes concatenation.
Then
the supertree would be produced in a greedy manner by iterating over
the list $S$, and adding each rooted split to a growing rooted
supertree if that split is compatible with the supertree.
In other
words, the set $C$ is initialized to the empty set, and we then
iteratively consider each split in the list $\mathcal{S}$ and insert
it into $C$ if the resulting set remains jointly compatible.
The BUILD
algorithm \citep{AhoSSU1981} is used both to check compatibility, and
to construct a compatible tree from the final set $C$.

>>>>

The Open Tree of Life project \citep{HinchliffEtAl2015} seeks to build
a comprehensive supertree \citep[see][]{redelings2017supertree}
covering all of life.
The approach is to combine information in
published phylogenies with a comprehensive taxonomy that supplies
taxon names.
The supertree method has been used with the Open Tree of
Taxonomy \citep[OTT hereafter, see][]{rees2017automated}, but 
could be used in more general contexts.
The taxonomy is a crucial input for several
reasons: it covers a wide range of species, its list of names allows
for alignment of different phylogenetic estimates to each other, and
it provides names for clades.
Only a small proportion of known species
have been included in a phylogenetic analysis, thus a taxonomy is
important for achieving comprehensive coverage of known taxa.
Phylogenetic estimates collected from the published literature often
use different names for the same species.
Lists of synonyms in OTT
(and other taxonomies) allow data curators to ``align'' input
phylogenies to the taxonomy \citep[see][ for discussion of the
curation process that the Open Tree of Life project uses to align phylogenetic
estimates to OTT]{McTavishEtAt2015}.
This alignment is important for recognizing when two different
estimates refer to the same taxon.
Biologists are familiar with names
for ``higher'' taxa (taxa above the species rank).
Thus, naming clades
in the supertree makes the tree easier to navigate and use.

Some taxonomies indicate which taxa are thought to be the result of hybridization, 
but the supertree approach
described here does not consider any hybridization.
Thus the taxonomic
hierarchy can be converted to a tree.
While not all taxonomists
believe that named taxa should correspond to clades \citep[see, for
example,][]{HorandlS2010}, the priniciples of phylogenetic
classification are used by such a large majority of practicing
taxonomists, that we have chosen to treat the taxonomic tree that
mirrors the hierarchy of the taxonomy as an estimate of phylogenetic
relationships.
We refer to labels of taxonomy nodes as ``taxon
names'', and assume that taxon names and taxonomy nodes have a one-to-
one correspondence.
Note that the taxonomy may also contain out-degree
1 nodes.
These nodes correspond to monotypic taxa which contain a
single child of lower taxonomic rank.

@@

Here we describe a supertree method so that we can resolve taxonomic
uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa.
Developing this method required deciding on a set of
operational semantics to be used when interpreting the \emph{incertae
sedis} annotation.
These semantics affect the interpretation of the
phylogenetic statements being made by the input taxonomy and the rules
for applying taxonomic names to the final supertree.
In addition to
discussing how to interpret the \emph{incertae sedis} annotation,
we describe the algorithmic changes to the supertree pipeline that
were required to adequately represent the taxonomic uncertainty.
The
approach described here allows us to include thousands of new taxa in
our supertree analysis that were previously filtered out.
It will also
enable the Open Tree of Life project to include extinct taxa, since
many of these taxa are \emph{incertae sedis}.
This makes substantial
progress towards our goal of comprehensive inclusion of known taxa
into the Open Tree of Life summary tree.

@@

The supertree algorithm of \citet{redelings2017supertree} takes as
input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of leaf-labeled rooted input phylogenies and a single rooted taxonomy
tree $\taxonomy_{p}$ that is ranked below all of the input trees.
As
discussed in \citet{redelings2017supertree}, the preferred tree would
display as many of the highly ranked input splits as possible.
If
computational barriers were not an issue, the method would act by
first producing a ranked list of rooted splits:
$\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$ where
$S(T_{i})$ denotes a list of non-trivial rooted splits created from a
post-order traversal of tree $i$ and '$+$' denotes concatenation.
Then
the supertree would be produced in a greedy manner by iterating over
the list $S$, and adding each rooted split to a growing rooted
supertree if that split is compatible with the supertree.
In other
words, the set $C$ is initialized to the empty set, and we then
iteratively consider each split in the list $\mathcal{S}$ and insert
it into $C$ if the resulting set remains jointly compatible.
The BUILD
algorithm \citep{AhoSSU1981} is used both to check compatibility.
 and to construct many aspects of a compatible tree from the final set $C$.
Refer to \citet{redelings2017supertree} for details of the procedure, including
a description of an optimization from
  pruning taxa that do not occur in any phylogenetic estimates in pre-processing
  and then reattaching them to a backbone tree.
This optimization is not crucial to the current work, but it does
  cause some groups to attach more tipward than they would if the entire
  tree were constructed using BUILD.
